Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-30 Manuscript/manuscript.Rmd
edits

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites), across resolution. We found that for random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the cell-line specific transcription factor binding sites, and was consistent across resolutions (Figure 5; Supplementary Figure 4).

![**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](Figures/figure5_proportion_of_importances_gm12878.png)

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that models built simply on cell line specific TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 6; Supplementary Figure 5). Additionally, when compared to the optimal model identified from the ensemble framework, one that included all 3 sets of annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations, indicating that, at lower resolution, the inclusion of other types of annotations contributed to additional noise rather than signal.

![**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](Figures/figure6_comparing_sets_of_annotations_gm12878.png)

@@

![**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)

>>>>

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 7). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

![**The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](Figures/proportion_of_importances_v2.png)

![**Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](Figures/figure6_comparing_sets_of_annotations_gm12878.png)

@@

![**Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)
